---
title: "decorate: Local correlation analysis of DNA methylation"
subtitle: 'Data from DNA methylation data post mortem brain from [Jaffe, et al. (2016)](https://www.nature.com/articles/nn.4181) '
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.Date()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
vignette: >
  %\VignetteIndexEntry{decorate: Local correlation of DNA methylation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

<!--- 
# run analysis

cd /sc/orga/projects/psychencode/gabriel/decorate_analysis
module purge
module load udunits proj gdal geos pandoc openssl pandoc R
source ~/.bash_profile
export https_proxy=http://mgmt01.chimera.hpc.mssm.edu:3128/
export http_proxy=http://mgmt01.chimera.hpc.mssm.edu:3128/
export ftp_proxy=http://mgmt01.chimera.hpc.mssm.edu:3128/

export OMP_NUM_THREADS=1
export R_LIBS=$R_LIBS_USER:$R_LIBS
R --vanilla

rmarkdown::render("/hpc/users/hoffmg01/scripts/decorate_analysis/methyl_scz_decorate.Rmd", output_dir='./', intermediates_dir='./'); system("cat methyl_scz_decorate.html | grep -v 'Found more than one class' | grep -v 'PythonEmbedInR' > methyl_scz_decorate2.html")

--->

Code from https://github.com/andrewejaffe/devMeth450k/blob/f2d50c438ae558edc96e852463fd2cee3be7768e/05_sz_diffs.R

```{r load.packages, echo=TRUE, message=FALSE, results='hide'}
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(decorate))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(data.table)) 
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(BiocParallel))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(qvalue))

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=TRUE,
  warning=FALSE,
  message=TRUE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  cache.lazy = FALSE,
  dev = c("png", "pdf"), 
  fig.width=7, fig.height=7)

options(markdown.HTML.stylesheet = 'css/custom.css')

register(SnowParam(4, "SOCK", progressbar=TRUE))
```

```{r load.data}
response = readRDS("/sc/orga/projects/psychencode/gabriel/decorate_analysis/data/rdas/response.RDS")
metadata = readRDS("/sc/orga/projects/psychencode/gabriel/decorate_analysis/data/rdas/metadata.RDS")
featureLocation = readRDS("/sc/orga/projects/psychencode/gabriel/decorate_analysis/data/rdas/featureLocation.RDS")

# features must have with of at least 2
end(featureLocation) = end(featureLocation) + 1

# Disease must be factor
metadata$Dx = factor(metadata$Dx, c("Control", "Schizo"))
```

# compute residuals
```{r get.residuals}
## Get residuals for decorate
design = model.matrix(~ Dx + Age + Race + negControl_PC1 + negControl_PC2 + negControl_PC3 + negControl_PC4, metadata)

fit = lmFit( response, design)
fit = eBayes( fit )

# # residuals + Dx
# residValues = residuals( fit, response) + coef(fit)[,'DxSchizo']
residValues = residuals( fit, response)

topTable(fit, coef='DxSchizo')
```

```{r get.resid.2}
# dcmp = svd( t(scale(t(residValues))), nu=2, nv=5)

# ## Get residuals for decorate
# design = model.matrix(~ Dx + Age + Race + negControl_PC1 + negControl_PC2 + negControl_PC3 + negControl_PC4, metadata) #  + dcmp$v[,1:5]

# fit = lmFit( response, design)
# fit = eBayes( fit )

# # residuals + Dx
# # residValues = residuals( fit, response) + coef(fit)[,'DxSchizo']
# residValues = residuals( fit, response)
```




# Learn local correlation structure
```{r hclustering, message=FALSE, echo=TRUE, results="hide"} 
# Compute correlation and hierarchical clustering on controls
treeList = runOrderedClusteringGenome( residValues[,metadata$Dx=='Control'], featureLocation, method.corr="spearman" )       
```


# Evaluate correlation structure versus distance
```{r plot.corr.decay}
# decay in original data
treeListOriginal = runOrderedClusteringGenome( response[,metadata$Dx=='Control'], featureLocation, method.corr="spearman" ) 
dfDistOriginal = evaluateCorrDecay( treeListOriginal, featureLocation, "chr22" )
plotCorrDecay( dfDistOriginal, method="R", xlim=c(500, 1e6), outlierQuantile=1e-5 )
   
# decay in residuals
dfDist = evaluateCorrDecay( treeList, featureLocation, "chr22" )
plotCorrDecay( dfDist, method="R", xlim=c(500, 1e6), outlierQuantile=1e-5 )
```

# Create clusters and then
# Measure strength of correlation structure in each cluster
```{r scoreClusters, echo=TRUE, message=FALSE}
treeListClusters = createClusters( treeList, method='meanClusterSize', meanClusterSize=c( 10, 20, 30, 40, 50, 100, 200, 500) )

# get total number of clusters
n_clusters = countClusters( treeListClusters )

# score each cluster to only retain  
# clusters with strong correlation structure
clstScore = scoreClusters(treeList, treeListClusters )
```


```{r hist.LEF}
df_LEF = do.call('rbind', clstScore )
df_LEF$id = factor(df_LEF$id, sort(unique(df_LEF$id)))

ggplot(df_LEF, aes(LEF, color=id)) + geom_density() + theme_bw(17) + theme(aspect.ratio=1)
```

<!---
```{r LEF.null}
# Evaluate LEF when features are permuted
# clustScoreNull = runPermutedData(response[,metadata$Dx=='Control'], featureLocation, method.corr="spearman",  meanClusterSize=c( 10, 20, 30, 40, 50, 100, 200, 500) ) 
      
# df_LEF = do.call("rbind", clustScoreNull$clusterScores)
# df_LEF$id = factor(df_LEF$id, sort(unique(df_LEF$id)))
# ggplot(df_LEF, aes(LEF, color=id)) + geom_density() + theme_bw(17) + theme(aspect.ratio=1) + ggtitle("LEF for permuted data")
```
--->

# Filter clusters based on strength of correlation
```{r filterClusters, echo=TRUE, message=FALSE, results="hide"}
# retain clusters based on filterign criteria
# If lead eigen value fraction (LEF) > 40% then ke/ep clusters
# LEF is the fraction of variance explained by the first eigen-value
clustInclude = retainClusters( clstScore, "LEF", 0.05 )

# get retained clusters  
treeListClusters_filter = filterClusters( treeListClusters, clustInclude )

# Collapse similar clusters
treeListClusters_collapse = collapseClusters( treeListClusters_filter, featureLocation )    
```


# Test differential signal
```{r diffcorr.test, echo=TRUE, message=FALSE, results="hide"}
# get total number of clusters
n_clusters = countClusters( treeListClusters_collapse )

# Evaluate Differential Correlation between two subsets of data
ecdBox = evalDiffCorr( residValues, metadata$Dx, featureLocation, treeListClusters_collapse, npermute, method = "Box.permute", method.corr="spearman")

ecdSLE = evalDiffCorr( residValues, metadata$Dx, featureLocation, treeListClusters_collapse, npermute, method = "deltaSLE", method.corr="spearman")

# get summary of results
df = summary( ecdBox )
 
# print results
head(df)
```

```{r print.df, echo=FALSE}
head(df)
```

Combine results to merge properties of each cluster into a single data.frame
```{r combine.results, message=FALSE, echo=TRUE}
df_results = combineResults( ecdBox, clstScore, treeListClusters, featureLocation)
head(df_results)
```

Of the `r sum(!is.na(df_results$pValue))` clusters tested, `r sum(df_results$p.adjust < 0.05, na.rm=TRUE)` have a adjusted p-value < 0.05.  Also, pi1=`r 1-qvalue(df_results$pValue)$pi0`.

### Summary of cluster properties
```{r cluster properties, message=FALSE}
# Histogram of LEF
ggplot(df_results, aes(LEF, fill=id)) + geom_histogram(alpha=0.7) + theme_bw(17) + xlim(0,1) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Requested mean cluster size") + xlab("Lead eigenvalue fraction (LEF)") + ggtitle("Summarize LEF")

# Histogram of mean absolute correlation
ggplot(df_results, aes(mean_abs_corr, fill=id)) + geom_histogram(alpha=0.7) + theme_bw(17) + xlim(0,1) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Requested mean cluster size") + xlab("Mean absolute correlation") + ggtitle("Summarize absolute correlation")

# Boxplot of number of features per cluster
df_results$id = factor(df_results$id, sort(as.numeric(unique(df_results$id))))
ggplot(df_results, aes(id, N, fill=id)) + geom_boxplot() + theme_bw(17) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Feature per cluster") + xlab("Requested mean cluster size") + ylab("Number of features") + ggtitle("Summarize feature per cluster") + coord_flip()
```



<!---
### Examine top cluster
```{r plot.clusters.best, echo=TRUE, cache=FALSE, message=FALSE, results="hide"}   
# extract peak ID's from most significant cluster
peakIDs = getFeaturesInCluster( treeListClusters_collapse, df$chrom[1], df$cluster[1], df$id[1])
    
# get location of peaks in this cluster  
query = range(featureLocation[names(featureLocation) %in% peakIDs])

# expand window to include adjacent clusters
window = 1e5 
start(query) = start(query) - window
end(query) = end(query) + window

library(EnsDb.Hsapiens.v75)      
ensdb = EnsDb.Hsapiens.v75
   
# Plot peaks in the query region

locText = with( data.frame(query), paste0(seqnames, ':', format(start, big.mark=','), '-', format(end, big.mark=',')))

main = paste0(df$chrom[1], ': cluster ', df$cluster[1], '\n', locText)

plotDecorate( ensdb, treeList, treeListClusters_collapse, featureLocation, query, showGenes=TRUE) + ggtitle(main)
```
--->

### Compare correlation structure along genome
First, plot correlation structure for controls (metadata\$Disease==0).  Then, plot correlation structure for cases (metadata\$Disease==1).

The cluster located at `r locText` has a p-value of `r format(df_results$pValue[1], digits=4)` and a test statistic of `r format(df_results$stat[1], digits=3)`.  The negative sign indicates a loss of correlation in the test set (i.e. Disease==1) compared to the baseline correlation in controls (i.e. Dissease==0)

```{r plot2Decorate.query, fig.width=15, message=FALSE, results="hide", echo=FALSE}
library(EnsDb.Hsapiens.v75) 
detach('package:EnsDb.Hsapiens.v75', unload=TRUE)
library(EnsDb.Hsapiens.v75)      
ensdb = EnsDb.Hsapiens.v75

figlist = lapply(1:50, function(i){
        
  # extract peak ID's from most significant cluster
  peakIDs = getFeaturesInCluster( treeListClusters_collapse, df$chrom[i], df$cluster[i], df$id[i] )
 
  # get location of peaks in this cluster  
  query = range(featureLocation[names(featureLocation) %in% peakIDs])

  # expand window to include adjacent clusters
  window = 100000 
  start(query) = start(query) - window
  end(query) = end(query) + window
  pdf("test.pdf", width=15)

  fig1 = plotDecorate( ensdb, treeList, treeListClusters_collapse, featureLocation, query, data=residValues[,metadata$Dx=='Control'])

  fig2 = plotDecorate( ensdb, treeList, treeListClusters_collapse, featureLocation, query, data=residValues[,metadata$Dx=='Schizo'])
  plot_grid( fig1, fig2, ncol=2, labels=c('A: Contols (i.e. Disease==0)', 'B: Cases (i.e. Disease==1)') )  
  dev.off()


})
figlist 
```

### Pairwise scatter plots
```{r scatter, echo=TRUE, message=FALSE}
plotScatterPairs( residValues, peakIDs, metadata$Dx) + ggtitle(main)
```

### Compare top cluster between cases and controls
```{r plot.results, cache=FALSE, fig.width=12, fig.height=12}    
# plot comparison of correlation matrices for peaks in peakIDs
#  where data is subset by metadata$Disease
main = paste0(df$chrom[1], ': cluster ', df$cluster[1])     
plotCompareCorr( residValues, peakIDs, metadata$Dx) + ggtitle(main)    
```

# Examine differential accessability signal for these peaks
```{r DE_peaks, echo=TRUE}
topTable(fit, coef='DxSchizo', number=Inf)[peakIDs,]
```


```{r save.significant.clusters}
loc = '/sc/orga/projects/psychencode/gabriel/decorate_analysis/bed/'

# all features
rtracklayer::export.bed( featureLocation, paste0(loc, "methylSz_all.bed"))

# background - only features consider in tests
featureNames_background = getFeaturesInClusterList( treeListClusters_collapse, chrom=df_results$chrom, clustID=df_results$cluster, id=df_results$id)
featureNames_background = unique(unlist(featureNames_background))
rtracklayer::export.bed( featureLocation[featureNames_background], paste0(loc, "methylSz_background.bed"))

# get significant peaks
idx = which(df_results$p.adjust < 0.05)
featureNames_signif = getFeaturesInClusterList( treeListClusters_collapse, chrom=df_results$chrom[idx], clustID=df_results$cluster[idx], id=df_results$id[idx])
featureNames_signif = unique(unlist(featureNames_signif))
rtracklayer::export.bed( featureLocation[featureNames_signif], paste0(loc, "methylSz_signif.bed"))

# get significant peaks - UP
idx = which(df_results$p.adjust < 0.05 & df_results$stat > 0)
featureNames_signif = getFeaturesInClusterList( treeListClusters_collapse, chrom=df_results$chrom[idx], clustID=df_results$cluster[idx], id=df_results$id[idx])
featureNames_signif = unique(unlist(featureNames_signif))
rtracklayer::export.bed( featureLocation[featureNames_signif], paste0(loc, "methylSz_signif_up.bed"))

# get significant peaks - DOWN
idx = which(df_results$p.adjust < 0.05 & df_results$stat < 0)
featureNames_signif = getFeaturesInClusterList( treeListClusters_collapse, chrom=df_results$chrom[idx], clustID=df_results$cluster[idx], id=df_results$id[idx])
featureNames_signif = unique(unlist(featureNames_signif))
rtracklayer::export.bed( featureLocation[featureNames_signif], paste0(loc, "methylSz_signif_down.bed"))

```

















